---
title: "Data visualization note"
author: "Ling"
date: "10/5/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)
library(patchwork)
```
# Questions for HW

  - *no $ in HW*

# data visualization and explanatory data analysis (EDA) 1

- EDA: between loading data and formal analysis
  - visualization
  - data completeness and reliability
  - quantification of centrality and variability
  - initial evaluation of hypothesis
  - hypothesis generation
- visualization
  - to look at data
  - eye-balling? --> should be quickly to make a plot in EDA

## `ggplot`

- support: google and stackoverflow
- "trouble in making a plot" <-- trouble with *data tidiness*
- *GGPLOT ASSUMES THE DATA ARE TIDY!* 

- graph components
  - data
  - aesthetic mappings (x=, y=)
  - geoms
  - *facets
  - *scales
  - *statistics

# practices

Using the same datasets in data wrangling.

## e.g.1 NOAA weather data

```{r}
weather_df = # pull data from 3 variables
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate( 
    name = recode( #create a new variable "name"
      id,  # transform station id into location
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())

weather_df
```

## making plots

- scatterplot
  - min temp and max temp
  
```{r scatterplot}
ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()

# or

weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
    geom_point()

# or

plot_weather = 
  weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) 

plot_weather + geom_point()
```

- make an advanced scatterplot
```{r advanced scatter}

ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = 0.5) + # alpha --> add transparency
  geom_smooth(aes(color = name), se = F) +
  facet_grid(. ~ name) # show the plots differently

# or
ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + # if want the option global for the code
  geom_point(alpha = 0.5) + # alpha --> add transparency
  geom_smooth(se = F) +
  facet_grid(. ~ name) 

# to turn off color in some plots
ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(aes(color = NULL), se = F) 

weather_df %>%
  ggplot(aes(x = date, y = tmax, color = name)) +
    geom_smooth(se = F) +
    geom_point() +
    facet_grid(. ~ name)
# what if "se = T"?
weather_df %>%
  ggplot(aes(x = date, y = tmax, color = name)) +
    geom_smooth(se = T)
# actually the SEs are not helpful

# learning assessment

weather_df %>%
  filter(name == "CentralPark_NY") %>%
  mutate(
      tmin = tmax * (9 / 5) + 32,
      tmax = tmin * (9 / 5) + 32) %>%
  ggplot() +
    geom_point() +
    geom_smooth()

# hex: heatmap
# need hexbin package
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
    geom_hex()

```

- Univariate plots
```{r univariate plots}
weather_df %>%
  ggplot(aes(x = tmax, color = name)) +
    geom_histogram() +
    facet_grid( . ~ name)

# notice color vs. fill

weather_df %>%
  ggplot(aes(x = tmax, fill = name)) +
    geom_density(alpha = 0.4, adjust = 0.5, color = "blue") 

```

- try other plots
  - box plot
```{r boxplot}
ggplot(weather_df, aes(x = name, y = tmax)) +
  geom_boxplot()

```
  - violin plot
```{r}
ggplot(weather_df, aes(x = name, y = tmax)) +
  geom_violin(aes(fill = name), alpha = .5) +
  stat_summary( fun = "median", color = "blue")
```

  - ridge plot: a replacement for both boxplots and violin plots. in `ggridges` package. Helpful when observing distributions across multiple categories
  
```{r ridge plot}
ggplot(weather_df, aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = .85)
```

n.b. 

- saving and embedding plots

```{r set the plot size, global option}
knitr::opts_chunk$set(
  fig.width = 0.6, 
  fig.asp = 0.6,
  out.width = "90%" # 90% of the full width, leave some space
)
```

or 

```{r fig.width= 12}
ggplot(weather_df, aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = .85)

```

# visualization 2

## questions last class

- `aes(color = "blue")` vs. `color = "blue"`
  - anything in `aes` is a variable
  - `color` out of `aes` is to color the plot.

- update Mr. Trash Wheel dataset

## lec

- a good pic
  - show as much of the data as possible
  - avoid superfluous frills
  - "what's the audience want to see?"
  - facilitate comparisons
    - DO NOT do pie charts unless describing a pie
    - group in a sensible order
    - common axes
    - color to highlight groups
    
- advanced graph components and customizing the plot

## practice

1. NOAA weather data
```{r review the scatter plot made in visualization 1}

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)

# add labels
weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs( # labs: title, axes, caption("ref" on the lower right)
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  ) +
  scale_color_hue(
    name = "location",
    h = c(100,300))

weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15ºC", "0", "15"),
    limits = c(-20, 30)) + 
  scale_y_continuous(
    trans = "sqrt", 
    position = "right")

# change the hue
library(viridis)
ggp_temp_plot = 
  weather_df %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    caption = "Data from the rnoaa package"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE # cont color gradient vs. discrete color
  )

ggp_temp_plot

# themes
# background color, elements in the plot

ggp_temp_plot + 
  theme(legend.position = "bottom") # put the legend at the bottom. "none" is to remove the legend

# get rid of the gray background
ggp_temp_plot +
  theme_bw() # black and white

# to make the plot in excel style, use ggthemes::themes_excel

ggp_temp_plot +
  ggthemes::theme_excel() # DO NOT  use the excel 2003 theme
```

*the order of `theme_bw()` and `theme` really matters!*

- theme_bw() --> theme()
  - the elements made by `theme_bw()` will be overwritten
- theme() --> theme_bw()
  - the elements made by `theme()` will be overwritten
  
*setting options*
put it at the beginning

# visualization3: EDA

## explanatory data analysis (EDA)

- EDA includes
  - visualization
  - checks for data completeness and reliability
  - initial evaluation of hypothesis
  - hypothesis generation
  
- grouping info.
  - numerical summaries of data now, esp. w/in groups
  - `group_by()` how data is structured
    - beneath the data frame, not show directly
  - `summarize()` compute one-number summaries

- NOTE
  - the blur line between EDA and formal analysis
  - only run the analysis necessary
  
## practice

```{r}
weather_df %>%
  mutate(
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything())
```

`group_by()`

```{r}
weather_df %>%
  mutate(
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything()) %>%
  group_by(name, month)
```
Use `ungroup()` to remove groups.

counting things using `summarize()`
```{r}
weather_df %>%
    mutate(month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything()) %>%
    group_by(month) %>%
    summarize(n_obs = n())

weather_df %>%
    mutate(month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything()) %>%
    group_by(name,month) %>%
    summarize(n_obs = n())

weather_df %>%
    mutate(month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, everything()) %>%
  pull(month) %>%
  table
```

use `janitor::tabyl` to create 2x2 tables
more general summaries
```{r}
weather_df %>%
  mutate(month = lubridate::floor_date(date, unit = "month")) %>%
  group_by(month) %>%
  summarize(
    n_obs = n(),
    n_days = n_distinct(date),
    tmax_mean = mean(tmax, na.rm = T),
    tmax_sd = sd(tmax, na.rm = T)) %>%
  ggplot(aes(x = month, y = tmax_mean, color = name)) +
    geom_point()
```

grouped mutate
```{r}
weather_df %>%
  mutate(
    month = lubridate::floor_date(date, unit = "month"),
    mean_tmax = mean(tmax, na.rm = T)) %>%
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
    geom_point()

# versus

weather_df %>%
  group_by(name) %>%
  mutate(
    month = lubridate::floor_date(date, unit = "month"),  
    mean_tmax = mean(tmax, na.rm = T)) %>%
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
    geom_point()
```
`mean_rank`
```{r}
weather_df %>%
  mutate(
    month = lubridate::floor_date(date, unit = "month"),  
    tmax_rank = min_rank(desc(tmax))) #min_rank to rank variables, desc for in a descending order
```

window function
lagged variables: how big is the day-to-day difference?
```{r}
weather_df %>%
  mutate(
    month = lubridate::floor_date(date, unit = "month"),  
    lag_tmax = lag(tmax,n = 1)) # comparing T_d vs. T_(d-1)

# versus
weather_df %>%
  group_by(name) %>%
  mutate(
    month = lubridate::floor_date(date, unit = "month"),  
    lag_tmax = lag(tmax,n = 1)) 
# the first lag_tmax of Wakiki is na this time

weather_df %>%
  group_by(name) %>%
  mutate(
    month = lubridate::floor_date(date, unit = "month"),  
    lag_tmax = lag(tmax,n = 1),
    tmax_diff = tmax - lag_tmax) %>%
  summarize(diff_sd = sd(tmax_diff, na.rm = T))
```
summary a linear model

```{r}
weather_df %>%
  group_by(name) %>%
  summarize(cor_min_max = cor(tmin, tmax, use = "complete"))

# versus
weather_df %>%
  filter(name == "CentralPark_NY") %>%
  lm(tmax ~ tmin, data = .)

# versus
weather_df %>%
  group_by(name) %>%
  lm(tmax ~ tmin, data = .)
```





 