---
title: "Data visualization note"
author: "Ling"
date: "10/5/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)
```
# Questions for HW

  - *no $ in HW*

# data visualization and explanatory data analysis (EDA)

- EDA: between loading data and formal analysis
  - visualization
  - data completeness and reliability
  - quantification of centrality and variability
  - initial evaluation of hypothesis
  - hypothesis generation
- visualization
  - to look at data
  - eye-balling? --> should be quickly to make a plot in EDA

## `ggplot`

- support: google and stackoverflow
- "trouble in making a plot" <-- trouble with *data tidiness*

- graph components
  - data
  - aesthetic mappings (x=, y=)
  - geoms
  - *facets
  - *scales
  - *statistics

# practices

Using the same datasets in data wrangling.

## e.g.1 NOAA weather data

```{r}
weather_df = # pull data from 3 variables
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USC00519397", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2017-01-01",
    date_max = "2017-12-31") %>%
  mutate( 
    name = recode( #create a new variable "name"
      id,  # transform station id into location
      USW00094728 = "CentralPark_NY", 
      USC00519397 = "Waikiki_HA",
      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) %>%
  select(name, id, everything())

weather_df
```

## making plots

- scatterplot
  - min temp and max temp
  
```{r scatterplot}
ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()

# or

weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
    geom_point()

# or

plot_weather = 
  weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) 

plot_weather + geom_point()
```

- make an advanced scatterplot
```{r advanced scatter}

ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = 0.5) + # alpha --> add transparency
  geom_smooth(aes(color = name), se = F) +
  facet_grid(. ~ name) # show the plots differently

# or
ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + # if want the option global for the code
  geom_point(alpha = 0.5) + # alpha --> add transparency
  geom_smooth(se = F) +
  facet_grid(. ~ name) 

# to turn off color in some plots
ggplot(weather_df, aes(x = tmin, y = tmax, color = name)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(aes(color = NULL), se = F) 

weather_df %>%
  ggplot(aes(x = date, y = tmax, color = name)) +
    geom_smooth(se = F) +
    geom_point() +
    facet_grid(. ~ name)
# what if "se = T"?
weather_df %>%
  ggplot(aes(x = date, y = tmax, color = name)) +
    geom_smooth(se = T)
# actually the SEs are not helpful

# learning assessment

weather_df %>%
  filter(name == "CentralPark_NY") %>%
  mutate(
      tmin = tmax * (9 / 5) + 32,
      tmax = tmin * (9 / 5) + 32) %>%
  ggplot() +
    geom_point() +
    geom_smooth()

# hex: heatmap
# need hexbin package
weather_df %>%
  ggplot(aes(x = tmin, y = tmax)) +
    geom_hex()

```

- Univariate plots
```{r univariate plots}
weather_df %>%
  ggplot(aes(x = tmax, color = name)) +
    geom_histogram() +
    facet_grid( . ~ name)

# notice color vs. fill

weather_df %>%
  ggplot(aes(x = tmax, fill = name)) +
    geom_density(alpha = 0.4, adjust = 0.5, color = "blue") 

```

- try other plots
  - box plot
```{r boxplot}
ggplot(weather_df, aes(x = name, y = tmax)) +
  geom_boxplot()

```
  - violin plot
```{r}
ggplot(weather_df, aes(x = name, y = tmax)) +
  geom_violin(aes(fill = name), alpha = .5) +
  stat_summary( fun = "median", color = "blue")
```

  - ridge plot: a replacement for both boxplots and violin plots. in `ggridges` package. Helpful when observing distributions across multiple categories
  
```{r ridge plot}
ggplot(weather_df, aes(x = tmax, y = name)) + 
  geom_density_ridges(scale = .85)
```

n.b. 

- saving and embedding plots
  






